About p-adic stability in linear algebra

### 2014-12-19 examples/stability.tex
First draft over LU factorization
Need to be improved…
Lacks numerical experiments.

We refer to \cite[Appendix B]{caruso-roe-vaccon:14a} for the computation
of the derivative of the LU factorization.
\todo{Complete this subsection.}

>>>>

\todo{Improve this Subsection.}

In this Subsection, we apply our framework to the case of $LU$ factorization. The computation of the differential of this factorization has already been done in \cite[Appendix B]{caruso-roe-vaccon:14a}. It remains to determine under which condition Proposition \ref{prop:precision} applies.

\begin{deftn}
Let $n \in \mathbb{Z}_{\geq 0}$ and  $O_n \subset K^{n \times n}$ be the set of the matrices such that all principal minors are non-vanishing. Let $U_n$ (resp. $L_n$)  be the $K$-vector spaces of the upper-triangular (resp. the strictly lower-triangular) matrices in $K^{n \times n}.$ Let $L_n^u \subset K^{n \times n} $ be the set of lower-triangular matrices with $1$'s one the diagonal. We define $D : O_n \rightarrow L_n^u \times U_n$ to be the mapping sending a matrix $M$ to $(L,M)$ such that $LU$ is the LU factorization of $M.$
\end{deftn}

This example has been previously studied by X.Caruso in \cite{caruso:12a}, but without using differential precision.
We show here that the framework of Subsection \ref{ssec:boundLambdaf} applies directly.

\subsubsection*{Computation of the differential of $D$} 
We begin with some notations:
\begin{deftn}
We denote by $\textrm{low}$ (resp. $\textrm{up}$) the canonical projection $K^{n \times n} \rightarrow L_n$ (resp. $K^{n \times n} \rightarrow U_n$).
\end{deftn}

Then, an explicit formula for $D'$ has been computed in \cite[Appendix B]{caruso-roe-vaccon:14a} :
\begin{prop}
Let $(M,L,U) \in O_n \times L_n^u \times U_n$ be such that $M=LU.$ Then,

\begin{tabular}{cccc}
$D'(M) :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
 & $\mathrm{d}M$ & $ \mapsto$ & $(L \textrm{low}(L^{*-1} \mathrm{d}M U^{-1}), \textrm{up}(L^{*-1} \mathrm{dM} U^{-1}) U).$ \\ 
\end{tabular} 
\end{prop}

A right inverse of $D'(M)$ can be made explicit:
\begin{rem} \label{rem:rightinverseLU}
Let \begin{tabular}{cccc}
$E(M) :$ & $L_n \times U_n$  & $\rightarrow$ &  $K^{n \times n}$\\ 
 & $(A,B)$ & $ \mapsto$ & $AU+LB$. \\ 
\end{tabular}

Then $D'(M)  \circ E(M) = Id.$
\end{rem}

\subsubsection*{Applying Subsection \ref{ssec:boundLambdaf}}

In order to apply Subsection \ref{ssec:boundLambdaf}, $f$ will be a translation of $D,$ so as to ensure $g(0) \neq 0.$ Namely,  let $(M_0,L_{0},U_{0}) \in O_n \times L_n^u \times U_n$ be such that $M_0 = L_{0}U_{0},$ we define

\begin{tabular}{cccc}
$f :$ &  $K^{n \times n}$ & $\rightarrow$ & $L_n \times U_n$ \\ 
 & $M$ & $ \mapsto$ & $(L_{M+M_0}-L_{0}, U_{M+M_0}-U_{0}).$ \\ 
\end{tabular}

Then, $f(0)=0$ and \[f'(M) \cdot \mathrm{d}M = (L_{M+M_0} \textrm{low}(L_{M+M_0}^{-1} \mathrm{d}M U_{M+M_0}^{-1}), \textrm{up}(L_{M+M_0}^{-1} \mathrm{dM} U_{M+M_0}^{-1}) U).\]

We now proceed with the three steps of the framework of Subsection \ref{ssec:boundLambdaf}.

\textbf{The $C$ constant:}

\begin{lem}
For $C = \max (\Vert U_{0} \Vert,\Vert L_{0} \Vert ),$ $B_{K^{n \times n}}(1) \subset f'(0) \cdot B_{K^{n \times n}}(C).$ 
\end{lem}
\begin{proof}
Apply Remark \ref{rem:rightinverseLU}.
\end{proof}

\textbf{The differential equation:}

\begin{lem} Let $g=g_{M_0}$ be defined by:
\begin{align*}
g(A,B) \cdot X =& ( (L_0^*+A) \textrm{low} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right), \\
& \: \textrm{up} \left( (L_0^*+A)^{-1}X(U_0+B)^{-1} \right) (U_0+B) ).\\
\end{align*}
Then, \[f' = g \circ f.\]
\end{lem}

\textbf{Determining $\mu$ and $\nu$}

Let $\vert \: \vert $ be an ultrametric norm on $K$ and $\Vert \: \Vert$ be the corresponding subordinate matrix norm. To determine $\mu$ and $\nu,$ we first use the following lemma

\begin{lem}
Let $(U,V) \in K^{n \times n} \times K^{n \times n}.$ If $\Vert V \Vert < \Vert U \Vert$ then $\Vert U+V \Vert = \Vert U \Vert.$

If $U \in GL_n(K)$ and $\Vert U \Vert < \Vert U^{-1} \Vert^{-1},$ then $\Vert (U+V)^{-1} \Vert = \Vert U^{-1} \Vert.$

Finally $\Vert U^{-1} \Vert^{-1} \leq \Vert U \Vert.$
\end{lem}

This leads to the following upper-bound on the norm of $g(A,B)$.

\begin{prop}
Let $(A,B) \in L_n \times U_n.$ be such that $\Vert A \Vert < \Vert L_0^{*-1} \Vert $ and $\Vert B \Vert <\Vert U_0^{*-1} \Vert. $ Then 
\[ \Vert g(A,B) \cdot X \Vert \leq \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right) \times \Vert X \Vert . \]
\end{prop}

\begin{prop}
For $f$ and $g$ as before, we can take $\nu = \min ( \Vert L_0^{-1} \Vert ^{-1} ,\Vert U_0^{-1} \Vert ^{-1} ),$ and \[ \mu = \max \left( \Vert L_0^* \Vert \Vert L_0^{*-1} \Vert  \Vert U_0^{*-1} \Vert, \Vert U_0^* \Vert \Vert U_0^{*-1} \Vert  \Vert L_0^{*-1} \Vert \right).\]
\end{prop}

\textbf{Conclusion}

Consequently, we can apply Proposition \ref{prop:boundLambdaf2} and Proposition 3.12 of \cite{caruso-roe-vaccon:14a}:

\begin{prop}
For any $\rho \in (0,1],$ if $r<e^{ \nu -2(\alpha +\mu)+\log \left( \frac{\rho}{C} \right)},$ the equation \ref{eq:firstorder} for $f$ is satisfied.
\end{prop}

\begin{ex}
If $\Vert L_0\Vert = \Vert L_0^{-1}\Vert =\Vert U_0\Vert =  \Vert U_0^{-1}\Vert = \rho=1$ and $p=2,$ then Proposition \ref{prop:precision} can be applied with $r=2^{-4}.$
\end{ex}

\subsubsection*{Numerical experiments}

\todo{Add numerical experiments...}
